### 2017-07-04 reasoning.tex
Edits.

\begin{figure}
\centering
\begin{tabular}{l|l}
\begin{txnimpcode}
$\begin{decoration}
  \{X = k\}
\end{decoration}$
  atomic {
    v1 := 0; v2 := 1;
    while(v1<X) {
      v2 := v2 * (X-v1);
      v1 := v1+1;
    }
    X := v2; Y := v2;
  }
$\begin{decoration}
  \{X = k! \conj Y = X\}
\end{decoration}$
\end{txnimpcode}
&
\begin{txnimpcode}
$\begin{decoration}
  \{X = k\}
\end{decoration}$
  transaction {
    v1 := 0; v2 := 1;
    while(v1<X) {
      v2 := v2 * (X-v1);
      v2 := v2 * v1;
      v1 := v1+1;
    }
    X := v2; Y := v2;
  }
$\begin{decoration}
  \{Y = X\}
\end{decoration}$
\end{txnimpcode}
\end{tabular}
  \caption{Factorial program written as an atomic block and a weakly
  isolated transaction (shared variables have uppercase names).
  Effect of multiple iteratons of the \C{atomic} loop can be
  summarized as $k!$, while same cannot be done with the
  \C{transaction} loop, where weak isolation allows reads in different
  iterations to return different values for $X$.  Nonetheless,
  post-state satisfies $X=Y$ in both cases due to atomicity of
  writes.}
\label{fig:atomic-vs-transaction}
\end{figure}

We now describe a proof system that lets us prove the correctness of a \txnimp
program $c$ w.r.t its high-level invariant $I$, on a machine that
satisfies the isolation specifications ($\I$) of its
transactions\footnote{Note the difference between $I$ and $\I$. The
former constitute \emph{proof} \emph{obligations} for the programmer,
whereas the latter describes a transaction's \emph{assumptions} about
operational characteristics of the underlying machine.}.  Our proof
system is essentially an adapatation of the rely-guarantee reasoning
framework to the setting of weakly isolated database transactions. It
deals with the additional complexity introduced by this setting, while
also taking advantage of the opportunities it provides to simplify the
reasoning. The primary challenge is the weak isolation; how do we
relate a transaction's isolation specification ($\I$) to its rely
relation ($R$) that describes the environment, so that the
interference is considered only insofar as the isolation level allows
it? Our proof system shows how to compute a rely relation ($R$) modulo
the isolation specification ($\I$) of a transaction. Another
characteristic of the transaction setting is the atomicity of a
trasaction's aggregate writes; a transaction's writes are not visible
to its peers until it commits. In the context of rely-guarantee, this
means that the transactions guarantee ($G$) should capture the
aggregate effect of a transaction, and not its individual writes.
While \C{atomic} blocks are also present in the shared-memory
programs, the fact that a transactions are weakly isolated introduces
complexity.  Unlike an \C{atomic} block, the effect of a transaction
is \emph{not} a sequential composition of the effects of its
statements because each statement witnesses a potentially different
version of the state. This is illustrated in
Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
\C{X} can be composed to describe the effect of an \C{atomic} block,
while same cannot be done in a transaction. Thus, weak isolation needs
to be taken into account when defining and verifying guarantees, and
our proof system shows how.

@@

Fig.~\ref{fig:rg-rules} shows an illustrative subset of the
rely-guarantee (RG) reasoning rules for $\txnimp$. We define two RG
judgments: top-level ($\rg{I,R}c{G,I}$), and transaction-local ($\R
\vdash \hoare{P}c{Q}$).  Recall that the standard RG judgment is the
quintuple $\rg{P,R}{c}{G,Q}$. Instead of a separate $P$ and $Q$, our
top-level judgment has $I$ for both pre- and post-conditions, because
our focus is on verifying that a \txnimp program \emph{preserves}
database's consistency conditions\footnote{The terms \emph{consistency
condition}, \emph{high-level invaraint}, and \emph{integrity
constraint} all mean the same.}. Transaction-local RG judgment doesn't
include a guarantee relation because transaction-local effects are not
visible outside a transaction. Also, the rely relation ($\R$) of the
transaction-local judgment is not same as the top-level rely relation
($R$) as it takes into account the transaction's isolation
specification ($\I$). Intuitively, $\R$ is $R$ modulo $\I$; formal
definition will be given shortly. Recall that a transaction writes to
its local database ($\stl$), which is then flushed when the
transaction commits. Thus, the guarantee of a transaction depends on
the state of its local database at the commit point. The pre- and
post-condition assertions ($P$ and $Q$) in the local judgment
facilitate tracking the changes to the transaction-local state, which
eventually helps us prove the validity of the transaction's guarantee.
Note that $P$ and $Q$ are bi-state assertions; they relate
transaction-local database state ($\stl$) to the global database state
($\stg$). Thus, the transaction-local judgment effectively tracks how
transaction-local and global states change in relation to each other.

The quintessential aspect of a rely-guarantee judgment is the
stability condition, which, intuitively, requires the validity of an
assertion $\phi$ to be unaffacted by the interference, i.e., the rely
relation $R$. In conventional RG, stability is defined as following
($\sigma$ denotes a state):
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and isolation
specification, we use multiple definitions of stability in
Fig.~\ref{fig:rg-rules}, but they all convey the same intuition as
above. We will introduce them as they are encountered.

@@

Recall that $P$ and $Q$ of the transaction-local RG judgment are
binary assertions; they relate local and global database states. The
local judgment rules require one or both of them to be stable. The
stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
is defined as following:
\begin{smathpar}
\begin{array}{c}
\forall \stl,\stg,\stg'.~ Q(\stl,\stg) \conj \R(\stl,\stg,\stg')
\Rightarrow Q(\stl,\stg')
\end{array}
\end{smathpar}
That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
by the isolation specfication (which implicitly considers the local
state $\stl$) takes $\stg$ to $\stg'$, the $Q$ must also relate $\stl$
to $\stg'$.

@@

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to the the inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of a
$\I$-constrained relation. We therefore introduce a side-condition on
$\I$ that restores the equality. The condition requires $\I$ to allow
an intereference $R^{*}(\stg,\stg'')$, for two database states $\stg$
and $\stg''$, only if it also allows interference for every prefix of
$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
shouldn't make the interference from $\stg$ to $\stg''$ valid. We call
this the stability condition on $\I$, defined as below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid, then an interference has already modified a
record, and no further interference will restore the original record,
because original record bears the id of a transaction that has long
committed. Thus, $\I_{ss}$ remains invalid.

>>>>

\begin{figure}
\centering
\begin{tabular}{l|l}
\begin{txnimpcode}
$\begin{decoration}
  \{X = k\}
\end{decoration}$
  atomic {
    v1 := 0; v2 := 1;
    while(v1<X) {
      v2 := v2 * (X-v1);
      v1 := v1+1;
    }
    X := v2; Y := v2;
  }
$\begin{decoration}
  \{X = k! \conj Y = X\}
\end{decoration}$
\end{txnimpcode}
&
\begin{txnimpcode}
$\begin{decoration}
  \{X = k\}
\end{decoration}$
  transaction {
    v1 := 0; v2 := 1;
    while(v1<X) {
      v2 := v2 * (X-v1);
      v2 := v2 * v1;
      v1 := v1+1;
    }
    X := v2; Y := v2;
  }
$\begin{decoration}
  \{Y = X\}
\end{decoration}$
\end{txnimpcode}
\end{tabular}
  \caption{Factorial program written as an atomic block and a weakly
  isolated transaction (shared variables have uppercase names).
  Effect of multiple iterations of the \C{atomic} loop can be
  summarized as $k!$, while same cannot be done with the
  \C{transaction} loop, where weak isolation allows reads in different
  iterations to return different values for $X$.  Nonetheless,
  post-state satisfies $X=Y$ in both cases due to atomicity of
  writes.}
\label{fig:atomic-vs-transaction}
\end{figure}

We now describe a proof system that lets us prove the correctness of a
\txnimp program $c$ w.r.t its high-level invariant $I$, on a machine
that satisfies the isolation specifications ($\I$) of its
transactions\footnote{Note the difference between $I$ and $\I$. The
former constitute \emph{proof} \emph{obligations} for the programmer,
whereas the latter describes a transaction's \emph{assumptions} about
operational characteristics of the underlying machine.}.  Our proof
system is essentially an adaptation of the rely-guarantee reasoning
framework to the setting of weakly isolated database transactions. It
deals with the additional complexity introduced by this setting, while
also taking advantage of the opportunities it provides to simplify the
reasoning. The primary challenge is the weak isolation; how do we
relate a transaction's isolation specification ($\I$) to its rely
relation ($R$) that describes the transaction's environment, so that
the interference is considered only insofar as the isolation level
allows it? Our proof system shows how to compute a rely relation ($R$)
modulo the isolation specification ($\I$) of a transaction. Another
characteristic of the transaction setting is atomicity; a
transaction's writes are not visible to its peers until it commits. In
the context of rely-guarantee, this means that the transactions
guarantee ($G$) should capture the aggregate effect of a transaction,
and not its individual writes.  While shared memory \C{atomic} blocks
also have the same characteristic, the fact that a transactions are
weakly isolated introduces complexity.  Unlike an \C{atomic} block,
the effect of a transaction is \emph{not} a sequential composition of
the effects of its statements because each statement witnesses a
potentially different version of the state. This is illustrated in
Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
a shared variable (\C{X}) can be composed to describe the effect of an
\C{atomic} block, while same cannot be done in a transaction. Thus,
weak isolation needs to be taken into account when defining and
verifying guarantees, and our proof system shows how.

@@

Fig.~\ref{fig:rg-rules} shows an illustrative subset of the
rely-guarantee (RG) reasoning rules for $\txnimp$. We define two RG
judgments: top-level ($\rg{I,R}c{G,I}$), and transaction-local ($\R
\vdash \hoare{P}c{Q}$).  Recall that the standard RG judgment is the
quintuple $\rg{P,R}{c}{G,Q}$. Instead of a separate $P$ and $Q$, our
top-level judgment has $I$ for both pre- and post-conditions, because
our focus is on verifying that a \txnimp program \emph{preserves}
database's consistency conditions\footnote{The terms \emph{consistency
condition}, \emph{high-level invariant}, and \emph{integrity
constraint} all mean the same.}. Transaction-local RG judgment doesn't
include a guarantee relation because transaction-local effects are not
visible outside a transaction. Also, the rely relation ($\R$) of the
transaction-local judgment is not same as the top-level rely relation
($R$) as it takes into account the transaction's isolation
specification ($\I$). Intuitively, $\R$ is $R$ modulo $\I$; formal
definition will be given shortly. Recall that a transaction writes to
its local database ($\stl$), which is then flushed when the
transaction commits. Thus, the guarantee of a transaction depends on
the state of its local database at the commit point. The pre- and
post-condition assertions ($P$ and $Q$) in the local judgment
facilitate tracking the changes to the transaction-local state, which
eventually helps us prove the validity of the transaction's guarantee.
Note that $P$ and $Q$ are bi-state assertions; they relate
transaction-local database state ($\stl$) to the global database state
($\stg$). Thus, the transaction-local judgment effectively tracks how
transaction-local and global states change in relation to each other.

The quintessential aspect of a rely-guarantee judgment is the
stability condition, which, intuitively, requires the validity of an
assertion $\phi$ to be unaffected by the interference, i.e., the rely
relation $R$. In conventional RG, stability is defined as following
($\sigma$ denotes a state):
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and isolation
specification, we use multiple definitions of stability in
Fig.~\ref{fig:rg-rules}, but they all convey the same intuition as
above. We will introduce them as they are encountered.

@@

Recall that $P$ and $Q$ of the transaction-local RG judgment are
binary assertions; they relate local and global database states. The
local judgment rules require one or both of them to be stable. The
stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
is defined as following:
\begin{smathpar}
\begin{array}{c}
\forall \stl,\stg,\stg'.~ Q(\stl,\stg) \conj \R(\stl,\stg,\stg')
\Rightarrow Q(\stl,\stg')
\end{array}
\end{smathpar}
That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
by the isolation specification (which implicitly considers the local
state $\stl$) takes $\stg$ to $\stg'$, the $Q$ must also relate $\stl$
to $\stg'$.

@@

A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to the the inductive reasoning: if
$Q$ is preserved by every step of $R$, then $Q$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of a
$\I$-constrained relation. We therefore introduce a side-condition on
$\I$ that restores the equality. The condition requires $\I$ to allow
an interference $R^{*}(\stg,\stg'')$, for two database states $\stg$
and $\stg''$, only if it also allows interference for every prefix of
$R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows interference
from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to $\stg''$
shouldn't make the interference from $\stg$ to $\stg''$ valid. We call
this the stability condition on $\I$, defined as below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid, then an interference has already modified a
record, and no further interference will restore the original record,
because original record bears the id of a transaction that has long
committed. Thus, $\I_{ss}$ remains invalid.
